Asymmetric relationships and their effects on coexistence: Supporting information

### 2023-07-31 paper/RefereeResponses1.tex

\noindent C35) Overall, I agree with the authors that the role of ATAs in coexistence has not been properly considered and the manuscript give convincing results of the importance of ATAs. \\

\noindent ***\emph{Thank you for the positive feedback!} \\
%DONE

\noindent C36) However in order to reach a broad audience of ecologists, the manuscript needs to reach a balance between mathematical definitions and simple explanations. Otherwise, it can remain to technical and of little importance to ecologists interested in the role of environmental variability of community dynamics. I will give some examples where the message can be improved, but I would suggest the authors to do the same effort themselves in other parts of the manuscript. \\

\noindent ***\emph{We have gone to great lengths to address the referee's specific suggestions below, and we have also redoubled our
efforts to make the manuscript as accessible as possible, generally. We appreciate being spurred to do so, thanks. } \\
%We have to redouble our efforts, as it says. Maybe we also want to point out somewhere that this particular
%sub-field is already a very technical one, but is nevertheless considered a mainstream and very important
%area in ecology? %food webs? networks

%I wrote the following, but decided not to include it here. Might include it elsewhere:
%The research field
%surrounding the MCT and mechanisms of coexistence is very important within ecology, but is nevertheless technical - see, for instance, the
%papers of Chesson, which have had enormous positive impact on the field, as well as the recent papers of Ellner which helped inspire
%our paper. All those papers, and many other papers in the field, are unavoidably technical, but have frequently been published
%in mainstream journals (the Ellner papers are in Ecology Letters) because of their importance and the importance of the subject
%matter they investigate. Though we agree with all efforts to make research results as clear as possible to non-technical audiences,
%we beg the referee

@@

\noindent C37) At the beginning of the introduction the authors cite previous work to show the prevalence and importance of ATAs for population dynamics, and the stability of ecological communities, and Fig.1 illustrate what ATAs mean in a clear example, yet it is not well explained or at least it was not clear to me how ATA can either promote or hinder coexistence in these initial sentences. \\

\noindent ***\emph{The referee appears to want us to present an intuitive explanation for how ATA effects work, and that 
is certainly idea. We have gone to great lengths to do so, and think we have succeeded. Though the referee's comment above may 
be interpreted to suggest that they wanted us
to add such an explanation to first paragraph of the paper, the reader is not equiped to understand the explanation at that 
point, so we added a new paragraph later in the Introduction (it is now the
second-to-last paragraph in the Introduction) that begins establishing such an intuition. We also added a new paragraph to the 
Discussion (the second paragraph of the Discussion). These paragraphs are also supported by two new SI sections and two new SI figures which 
explain, in detail, how the intuition established in the two new main text paragraphs emerges from the modelling (SI sections S11 
and S12, SI Figs S4 and S5).
%DAN: make sure these are the final numbers of these SI sections
We believe that intuition such as the new material establishes should always be based in formal theoretical and 
empirical work, so the focus of the paper remains on our expansion of the MCT and our results showing
the *importance* of ATA effects. Though we definitely think our additions giving an intuition for ATA effects are a large
improvement in the manuscript, so we thank the referee for spurring us to make these additions.  } \\
%DONE except the checking as noted above.

\noindent C38) Then the explanation of the importance of ATAs is done in relationship with the explanation of the storage effect. In detail, the explanation of the storage effect is superb, ... \\

@@

\noindent ***\emph{Thank you for the positive feedback!} \\
%DONE

\noindent C39) ...but how ATAs enter in play is not well solved. It is just mentioned that ATAs can modify the EC covariation, which I believe is not enough, a reader might not understand later the explanation when the lottery model is presented. \\

\noindent ***\emph{See our response to C37. We have added substantial material to the Introduction and Discussion that adds an
intuitive explanation for ATAs, and we thank the referee for spurring us to do so. 
We find it unlikely that any focussed reader will have substantial difficulties understanding the lottery model
material in the Introduction, since the lottery model is certainly no more complicated than models such as the Lotka-Volterra 
competition and predator-prey models, which are covered in essentially every introductory undergraduate ecology class, and we can assume
that almost all readers of Ecology Letters have taken such a class. See also comment C25 above, which indicates that another
referee found this material to be compelling.} \\
%DONE

@@

\noindent C40) ATAs can not only modify the importance of the storage effect but also of other mechanisms such as relative non-linearity, which is based on how competition fluctuates in ways that favor the competitor. Could the authors briefly indicate how this could occur? \\

\noindent ***\emph{This cannot occur. ATAs can only influence storage effects, not the other mechanisms, as stated at
the top of section 2.4 and explained mathematically in the SI section cited there. Basically, as explained in 2.4, 
this is because ATAs are about covariation between $E$ and $C$, whereas the mechanisms other than storage effects
do not pertain at all to covariation. For instance, relative nonlinearity is about the effects of variation 
in $E$ or $C$, not covariation between them.} \\
%DONE

\noindent C41) The lottery model is used to illustrate the importance of ATAs, and two types of fecundities are taken 
into account (log-normal and beta), can the author explain why both are important, and later on the results and discussion 
why ATAs play a major role with beta fecundities and not with log-normal.
The model used for the chemostat is in its formulation quite different compared to the lottery model. Although the technical
details are left for the supplementary material, can the authors explain more in the detail in the main text how the 
importance of ATAs can be used for this particular model.\\

\noindent ***\emph{This is certainly a valid question, and we have gone to some lengths to address it. 
Partly in response to this comment and partly in response to earlier comments of referee 3, 
we added substantial new material providing intuitive explanations of ATA effects in the Introduction,
Discussion, and SI (see above responses). That new material also has information comparing the models
and explaining why ATA effects are substantial for the beta-fecundities lottery model and not as strong
for the log-normal fecundities lottery model. We believe additional explanations and intuitive 
explorations could and should be the subject of several future papers, but we've added material to this manuscript
that takes that process as far as possible within the scope of the current paper, given length limits and
requests from the referees to simplify.} \\
%DONE

\noindent C42) The chemostat model is parameterized with information from a previous study, but does this model reproduce observed dynamics of the diatom system? this needs to be shown otherwise the importance of ATAs will be considered for another dynamics that do not correspond to the observed ones. \\

### 2023-07-31 paper/Paper.Rnw
working on brevity and clarity throughout. also added Jasmin's water limitation example to the Intro

%Intro the paper by intro'ing ATAs
At the core of ecology is the study of relationships between organisms and their environments.
Ecologists may ask, for instance, how local environment relates to 
population growth rate. To study such relationships, ecologists 
very commonly use metrics such as correlation and covariance. But these standard measures of the strength 
and direction of association between variables often do not tell the whole story 
\citep{Ghosh_2020_AER}.
To illustrate, the three pairs of variables in Fig. \ref{fig:pedagog1}a-c all have the same 
Pearson correlation despite striking differences in association structure: Fig. \ref{fig:pedagog1}b shows a 
symmetric type of association, while Fig. \ref{fig:pedagog1}a,c exhibit what has previously been called 
``asymmetric tail association'' [ATA; \citealp{Ghosh_2020_AER}]. 
Metrics of association other than correlations and covariance are also often used (e.g., various forms of 
regression), but correlations are common, and other common methods also reveal only a small portion 
of the information content of the relationship between variables \citep{Nelsen2006}.
ATAs and related ideas had previously been seldom considered in ecology, but recently several 
insights have been gained by studying how features of associations beyond standard measures 
impact various ecological phenomena \citep{deValpine_2014, Anderson_2019, Popovic_2019, Ghosh_2020_AER, Ghosh_2020_Eco,
Ghosh_2020_ecolevol, Ghosh_2021}. For instance, \cite{Ghosh_2020_Eco} argued that extinction risks can become elevated 
when metapopulations are subject to ATAs, e.g., if environmental ``catastrophes'' are more spatially synchronous 
across component populations
than are years of good environment, limiting the potential for rescue effects. 
And \cite{Ghosh_2021} showed that ATAs can alter the stability of 
ecosystem functioning. Fig. \ref{fig:pedagog1}d,e show contrasting examples of ATAs in nature using 
plankton population density time series; many other ecological examples of ATAs have also been explored 
\citep{Ghosh_2020_AER,Ghosh_2020_ecolevol,Ghosh_2021}. To our knowledge, the implications of ATAs for 
competition and coexistence have not been investigated. Doing so is the goal of this study.

%Intro to competition, coexistence, MCT, etc.
Competition, coexistence, and the maintenance of biodiversity are research topics at the core of modern ecology 
\citep{MacArthur_1958, Hutchinson_1961, Schmida_1984}. The competitive exclusion principle states 
that two species competing for the same limiting resource cannot coexist at constant densities - one will 
instead outcompete the other and drive it extinct \citep{Hardin_1960}.
Modern coexistence theory [MCT; \cite{Chesson_1994,Chesson_2000,Barabas_2018}] is a 
framework for understanding how species nevertheless coexist, and biodiversity is maintained, in real ecosystems,
through the niche partitioning effects of environmental variation.
MCT breaks down coexistence into the contributions of various mechanisms, with names such as ``storage effects'' \citep{Chesson_1994}; storage effects are reviewed conceptually below and will be
defined formally in the Theory section. MCT has been
applied to several systems, including annual \citep{Angert_2009, Hallett_2019} and perennial plants \citep{Adler_2006, Adler_2010}, 
tropical trees \citep{Usinowicz_2012}, plankton \citep{Caceres_1997, Descamps_2005, Letten2018}, and corals \citep{Alvarez-Noriega_2020}. 
%Narawani_2013
Although MCT is somewhat mathematically complex, recent extensions make it more accessible \citep{Ellner_2016,Ellner_2019}.

%Intro to storage effects and why they might be influenced by ATAs
It is possible to see why it is plausible that
ATAs may influence one mechanism of coexistence in particular, storage effects. We review, conceptually, 
what storage effects are \citep{Chesson_1994,Chesson_2012,Ellner_2016,Barabas_2018}, and explain why ATAs 
may be relevant to them. Storage effects allow competing species to fluctuate in abundance, without any going extinct,
by providing a relative benefit to whichever species is currently rare \citep{Chesson_2000}.
Storage effects hinge on covariation between the fluctuating quality of the environment
for a species (E) and the strength of competition (C) experienced by that species (called EC covariance);
and on the assumption that environmental responses for distinct species are also distinct,
so that no two species respond in the same way to environmental variation. Then, good environmental
conditions for a currently common species are paired with strong intraspecific competition, because of
positive EC covariance and the commonness of the species. But good environmental conditions for 
a currently rare species are not as closely accompanied by strong competition because the rarity of
the species limits competitive influence. The resulting asymmetry is what provides the relative benefit to the
rare species. Storage effects can be quantified by measuring to what extent EC covariance contributes to 
differences between currently rare and currently common species growth rates \citep{Chesson_1994,Ellner_2016}.  
ATAs may, in principle, influence storage effects simply because storage effects hinge on EC covariation, 
and ATAs, when they occur between environment and competition variables, can be an aspect of EC covariation.

%Lottery model and results based on it
To further demonstrate the plausibility of ATAs altering coexistence between species, we carried out 
simple simulations using a two-species model, the well-known lottery model. The simulations described in this
paragraph are sufficient to demonstrate that ATAs influence coexistence in some manner; the
precise nature of that influence is the subject of the rest of the paper.
The lottery model, which was originally introduced as a model of coral reef fish competing for a set of $N$ breeding territories, 
is as follows.
Letting $N_i(t)$ denote the adult population density of species $i=1,2$ at time $t$, and defining $N = N_1(t)+N_2(t)$,
model equations are
\begin{equation}
N_i(t+1)=(1-\delta)N_i(t)+\delta N \frac{B_i(t)N_i(t)}{B_1(t)N_1(t)+B_2(t)N_2(t)}  \label{model_eq}
\end{equation}
for $i=1,2$. Here, $\delta$ is an adult mortality rate, and $B_i(t)$ is the fecundity of species $i$ at time $t$. 
The model postulates that adults die at rate $\delta$ at each time step, and are replaced by juveniles 
in proportion to the reproductive outputs of the two species that year. Note that $N = N_1(t)+N_2(t)$ 
is constant through time. 
We assume for simplicity that the random variables $(B_1(t),B_2(t))$ are independent and identically 
distributed (iid) through time. In three distinct simulations (Fig. \ref{fig:pedagog1}f-h), 
$B_1$ and $B_2$ were, respectively, 
left- (Fig. \ref{fig:pedagog1}a), right- (Fig. \ref{fig:pedagog1}c) and symmetrically (Fig. \ref{fig:pedagog1}b)
tail associated, while the same marginal distributions for these fecundities were used in all 
simulations (see Methods
for details). Thus, the model setup was identical in the three simulations except for different ATAs. 
ATAs substantially influenced aspects of species coexistence in these simulations (Fig. \ref{fig:pedagog1}f-h).
Notation for the whole paper is summarized in Table \ref{tab:notation}.
Typesetting of mathematics by ecological journals is sub-optimal,
so we appended a latex version of this main text to the end of the 
Supporting Information (SI). Readers interested in the mathematical aspects should read that version.
%***DAN: Don't forget to do this

%Some intuition
We next give some initial intuition about how ATA effects on coexistence work, 
providing additional intuition in the Discussion.
It is known that storage effects can occur if a buffering mechanism reduces population 
losses in bad years so that
gains in good years can more than make up for them. 
Bad years happen when the negative effects of $C$ outweigh any 
positive effects of $E$; but this can occur
both 1) when $E$ is moderately favorable but $C$ is very strong, as well as 2) when $E$ is 
unfavorable and $C$ is not too weak.
Conversely, good years happen when positive effects of $E$ outweigh negative effects of $C$;
but this can occur 1) when $E$ is favorable and $C$ is not too strong,
as well as 2) when $E$ is moderately unfavorable but $C$ is very weak. 
It turns out that the potential for storage events to occur when 
$E$ and $C$ are both higher than average (the cases 1, above); versus the potential  
when $E$ and $C$ are
both lower than average (the cases 2) are not necessarily equal -- storage potential may differ
across the spectrum from 1 to 2.
ATA effects quantify the extent to which asymmetries of association between $E$ and 
$C$ are aligned with differences in the potential for storage across the spectrum from 1 to 2.
If $E$ and $C$ are largely unassociated in the parts of their distribution where
storage potential is high, ATA effects will be high. ATA effects quantify how temporal niche partitioning
varies across $EC$ spectrum spanned by the cases 1-2 sketched above.

%Goal para
We previously demonstrated that ATAs are common in nature and can influence ecological phenomena
\citep{Ghosh_2020_AER, Ghosh_2020_Eco, Ghosh_2020_ecolevol, Ghosh_2021}; and our arguments above
indicate that ATAs may influence species coexistence. Thus, we here seek to quantify the consequences 
of ATAs for competition, coexistence, and the maintenance of biodiversity, by means of the 
following specific goals. 
G1) We augment MCT to show formally how ATAs play a role in coexistence. 
MCT decomposes a measure of coexistence into components due to each of several mechanisms. 
We address how ATAs contribute to these components. 
%We examine mechanisms of coexistence that apply in models exhibiting solely temporal variation; 
%spatial variation can be considered in future work.
G2) We explore, using the simple lottery model, the relative magnitudes 
of the contributions to coexistence that come
from ATAs compared to other, previously known mechanisms of coexistence; and we determine whether species may 
sometimes coexist with ATAs but not without, or vice versa. This is to help 
illuminate whether ATA contributions to coexistence and biodiversity maintenance
are likely to be negligible or meaningfully large. 
G3) We provide an empirical example of species coexistence which highlights the role ATAs can play and the
potential importance of ATAs relative to other mechanisms of coexistence.
The example is a diatom system in a chemostat. In the Discussion, we consider whether climate change may alter
ATAs of environmental variables and thereby influence competition and coexistence.
Overall, our study presents a new mechanism of species coexistence
and a means of understanding its theoretical and empirical importance.

@@

We here pursue goal G1 from the Introduction: to augment MCT to show formally how ATAs play a role in coexistence.

@@

We develop our new theory for a general two-species model, later applying it to specific scenarios.
Our mathematical notation parallels
that of \cite{Ellner_2016,Ellner_2019}. It is assumed that the growth rate $r_i(t)=\ln(N_i(t+1)/N_i(t))$ 
of species $i$ can be written as an increasing function of
an environment-dependent factor $E_i(t)$ and as a decreasing function of a quantity representing
competitive pressure on species $i$, $C_i(t)$, i.e., $r_i(E_i,C_i)$. We also assume, for simplicity, 
that the $E_i(t)$ are iid through time, and likewise for the
$C_i(t)$. Though the iid assumption is probably not necessary, it simplifies the
presentation of new ideas.

@@

%ATA specialists
Contributions of ATAs to species coexistence open the possibility that
``ATA specialists'' may exist. For instance, if a strong competitor is present across a
collection of habitat patches on a landscape and the GWR of a weaker competitor is only positive 
in a patch if there are 
ATAs between the temporal fluctuations of two environmental variables in that patch,
then the weaker competitor can only persist in patches for which ATAs are present.
The weaker competitor can therefore perhaps be regarded as an ``ATA specialist'' in the landscape. EC-associations
can boil down to associations between two environmental variables, as was the case
for the lottery model. For the lottery model, $B_i(t)=E_i(t)$ was assumed, and $C_{i \bs i}$ simplified to $E_j(t)/\delta$ (see SI section 
\ref{SI-SIsect:TheoryForLotteryBoth}). Thus, EC-associations parallel associations between the two
environmental variables $E_i(t)$ and $E_j(t)$ for the lottery model, and ATA specialists may exist. 
Mechanisms of coexistence should correspond to
means by which niche differentiation is possible. Thinking about our new mechanism of 
coexistence in terms of possible ATA specialists helps highlight this connection. Likewise,
because we have described ATA exclusion as well as ATA rescue, it is possible to 
imagine a scenario in which a weaker competitor can only exist in habitat patches 
for which ATAs between environmental variables are absent, and is hence a specialist on scenarios with symmetric
tail associations of variables (an ``STA specialist''). Additional 
research will be needed to assess to what degree these possibilities are realized 
in nature.

%Climate change may alter ATAs of coexistence-relevant ATAs
An additional motivation for quantifying the importance of ATAs for species coexistence 
is that there is reason to believe climate change may
alter ATAs of environmental variables which impact coexistence, as follows.
It is well known that climate change can alter the intensity of extreme environmental events, including
heat waves, floods, hurricanes and fires \citep{Meehl_2004, Jentsch_2007, Ummenhofer_2017, Lyon2019, Keelings2020}.
If extreme events are not only becoming more intense, but are also increasing in duration (e.g., heat waves 
are not only hotter but also last longer, \citealp{Meehl_2004,Lyon2019,Keelings2020}), 
then the nature of the covariation between phenologically separated environmental signals
may be modified by these changes; in particular, ATAs between such variables may be altered.
For instance, imagine a case in which early spring temperature influences the growth of a plant species, A,
whereas late spring temperature influences plant species B. If these species are competitors, then EC covariation
contributing to their coexistence, or failure to coexist, could boil down to covariation
between early and late spring temperatures (e.g., if competitive dynamics are governed by the lottery model with $E_i$ representing
early spring temperature and $E_j$ representing late spring temperature, see previous paragraph). 
Because longer lasting heat waves are more likely to contribute
to both early and late spring temperature, a tendency for heat waves to become both more extreme and
longer lasting increases the degree of right-tail association between early and late spring temperature.
Likewise, a tendency for cold snaps to become both more intense (i.e., colder temperatures) and longer lasting can increase
left-tail association between early- and late-spring temperatures. As we showed in this study, changes in ATAs can then
result in different competitive outcomes between species.
To our knowledge, the potential importance of climatic changes in ATAs for species coexistence has
not been studied, though the importance for coexistence of changes
in means and variability of environmental variables has been studied in many systems
[e.g., \cite{White_2001, Adler_2006, Jentsch_2007, Angert_2009, diPaola_2012, Descombes_2020, Usinowicz_2021}].
Changes in variability and in extreme events are distinct from the changes in ATAs we
consider here, because ATAs pertain to \emph{relationships} between environmental and other
variables in the extremes, which are distinct from and can be altered independently
of changes in the extremes of the univariate marginal distributions themselves \citep{Nelsen2006,Ghosh_2020_AER}.

%Differences with the original, analytic version of the MCT
Our approach to exploring the influence of ATAs on coexistence relies on the computational
extension of MCT of \cite{Ellner_2019},
and our results help demonstrate strengths of that approach.
Storage effects as computed using the original, analytic approach to MCT 
\citep{Chesson_1994,Chesson_2000} should equal our term $\Delta^{[E \nat C]}$, 
which we refer to as the contributions to coexistence of correlation \emph{per se}.
This is because the weak-noise/Taylor approximation approach used in the original 
MCT neglects the effects of ATAs when higher-order terms
in Taylor expansions are dropped. The term $\Delta^{(E \shp C)}$, which our theory 
inherits from \cite{Ellner_2016}, is also dropped. These observations do not diminish 
the major strengths of the 
original, analytic approach, which are well known and also elaborated by \cite{Ellner_2019};
one such strength is the generality that analytic approaches provide.
It may be possible to expand the mathematics of the original MCT to consider ATAs.
Additional discussion points are in SI sections \ref{SI-sect:vital_rates}-\ref{SI-sect:storage_importance}.

>>>>

%Intro the paper by intro'ing ATAs
At the core of ecology is the study of relationships between organisms and their environments.
Ecologists may ask, for instance, how local environment relates to 
population growth rate. To study relationships, ecologists 
very commonly use metrics such as correlation. But standard measures of the strength 
and direction of association between variables do not tell the whole story 
\citep{Ghosh_2020_AER}.
To illustrate, the three pairs of variables in Fig. \ref{fig:pedagog1}a-c have the same 
Pearson correlation despite striking differences in association structure: Fig. \ref{fig:pedagog1}b shows a 
symmetric type of association, while Fig. \ref{fig:pedagog1}a,c exhibit what has been called 
``asymmetric tail association'' [ATA; \citealp{Ghosh_2020_AER}]. 
Metrics of association other than correlations are also often used (e.g., various forms of 
regression), but other common methods also reveal only a small portion 
of the information content of the relationship between variables \citep{Nelsen2006}.
ATAs and related ideas had previously been seldom considered in ecology, but recently 
insights have been gained by studying how features of associations beyond standard measures 
impact various ecological phenomena \citep{deValpine_2014, Anderson_2019, Popovic_2019, Ghosh_2020_AER, Ghosh_2020_Eco,
Ghosh_2020_ecolevol, Ghosh_2021}. For instance, \cite{Ghosh_2020_Eco} argued that extinction risks can become elevated 
when metapopulations are subject to ATAs, e.g., if environmental ``catastrophes'' are more spatially synchronous 
across component populations
than are years of good environment, limiting rescue effects. 
And \cite{Ghosh_2021} showed that ATAs can alter the stability of 
ecosystem functioning. Fig. \ref{fig:pedagog1}d,e show contrasting examples of ATAs in nature using 
plankton population density time series; many other examples of ATAs in nature have also been explored 
\citep{Ghosh_2020_AER,Ghosh_2020_ecolevol,Ghosh_2021}. To our knowledge, the implications of ATAs for 
competition have not been investigated. Doing so is our goal.

%Intro to competition, coexistence, MCT, etc.
Competition and coexistence are research topics at the core of modern ecology 
\citep{MacArthur_1958, Hutchinson_1961, Schmida_1984}. The competitive exclusion principle states 
that two species competing for the same limiting resource cannot coexist at constant densities - one will 
instead drive the other extinct \citep{Hardin_1960}.
Modern coexistence theory [MCT; \cite{Chesson_1994,Chesson_2000,Barabas_2018}] is a 
framework for understanding how species nevertheless coexist, and biodiversity is maintained, in real ecosystems,
through the niche partitioning effects of environmental variation.
MCT breaks down coexistence into the contributions of various mechanisms, with names such as ``storage effects'' \citep{Chesson_1994}; storage effects are reviewed conceptually below and 
defined formally in Theory. MCT has been
applied to several systems, including annual \citep{Angert_2009, Hallett_2019} and perennial plants \citep{Adler_2006, Adler_2010}, 
tropical trees \citep{Usinowicz_2012}, plankton \citep{Caceres_1997, Descamps_2005, Letten2018}, and corals \citep{Alvarez-Noriega_2020}. 
%Narawani_2013
Although MCT is somewhat mathematically complex, recent extensions make it more accessible \citep{Ellner_2016,Ellner_2019}.

%Intro to storage effects and why they might be influenced by ATAs
We indicate why it is plausible that
ATAs may influence one mechanism of coexistence in particular, storage effects, starting by conceptually reviewing 
what storage effects are \citep{Chesson_1994,Chesson_2012,Ellner_2016,Barabas_2018}. 
Storage effects allow competing species to fluctuate in abundance, without any going extinct,
by providing a relative benefit to whichever species is currently rare \citep{Chesson_2000}.
Storage effects hinge on covariation between the fluctuating quality of the environment
for a species ($E$) and the strength of competition ($C$) experienced by that species (called $EC$ covariance);
and on the assumption that no two species respond in the same way to environmental variation. Then, good environmental
conditions for a currently common species are paired with strong intraspecific competition, because of
positive $EC$ covariance and the commonness of the species. But good environmental conditions for 
a currently rare species are not as closely accompanied by strong competition because the rarity of
the species limits competitive influence. The resulting asymmetry is what provides the relative benefit to the
rare species. Storage effects can be quantified by measuring to what extent $EC$ covariance contributes to 
differences between growth rates of currently rare versus currently common species \citep{Chesson_1994,Ellner_2016}.  
ATAs may, in principle, influence storage effects simply because storage effects hinge on $EC$ covariation, 
and ATAs can be an aspect of $EC$ covariation. One scenario in which it is reasonable to imagine that
ATAs may be present in $EC$ covariation pertains to water limitation. Suppose two plants compete primarily for 
water, when water is scarce, but their niches become more differentiated when water is abundant.
Then the quality of the environment, $E$, as experienced by whichever plant is scarce, may only correspond closely to the 
competition, $C$, imposed on it by the other, currently abundant plant, when water is scarce and $E$ is low.

%Lottery model and results based on it
To further demonstrate the plausibility of ATAs altering coexistence, we carried out 
simple simulations using a two-species model, the lottery model. The simulations described in this
paragraph demonstrate that ATAs influence coexistence in some manner; the
nature of that influence is the subject of the rest of the paper.
The lottery model, which was introduced as a model of coral reef fish competing for a set of $N$ territories, 
is as follows.
Letting $N_i(t)$ denote the adult population density of species $i=1,2$ at time $t$, and defining $N = N_1(t)+N_2(t)$,
model equations are
\begin{equation}
N_i(t+1)=(1-\delta)N_i(t)+\delta N \frac{B_i(t)N_i(t)}{B_1(t)N_1(t)+B_2(t)N_2(t)}  \label{model_eq}
\end{equation}
for $i=1,2$. Here, $\delta$ is an adult mortality rate, and $B_i(t)$ is the fecundity of species $i$ at time $t$. 
The model postulates that adults die at rate $\delta$ at each time step, and are replaced by juveniles 
in proportion to the reproductive outputs of the two species that year. Note that $N = N_1(t)+N_2(t)$ 
is constant. 
We assume that the random variables $(B_1(t),B_2(t))$ are independent and identically 
distributed (iid) through time. In three distinct simulations (Fig. \ref{fig:pedagog1}f-h), 
$B_1$ and $B_2$ were, respectively, 
left- (Fig. \ref{fig:pedagog1}a), right- (Fig. \ref{fig:pedagog1}c) and symmetrically (Fig. \ref{fig:pedagog1}b)
tail associated, while the same marginal distributions were used in all 
simulations (see Methods
for details). Thus, the setup was identical in the three simulations except for different ATAs. 
ATAs substantially influenced aspects of species coexistence in these simulations (Fig. \ref{fig:pedagog1}f-h).
Notation for the paper is summarized in Table \ref{tab:notation}.
Typesetting of mathematics by ecological journals is sub-optimal,
so we appended a latex version of this main text to the end of the 
Supporting Information (SI); readers interested in the math should read that version.
%***DAN: Don't forget to do this

%Some intuition
We next give some initial intuition about how ATA effects on coexistence work, 
with more details in the Discussion.
It is known that storage effects can occur if a buffering mechanism reduces population 
losses in bad years so that
gains in good years can more than make up for them. 
Bad years happen when the negative effects of $C$ outweigh any 
positive effects of $E$; but this can occur
both 1) when $E$ is moderately favorable but $C$ is very strong, and 2) when $E$ is 
unfavorable and $C$ is not too weak.
Conversely, good years happen when positive effects of $E$ outweigh negative effects of $C$;
but this can occur 1) when $E$ is favorable and $C$ is not too strong,
and 2) when $E$ is moderately unfavorable but $C$ is very weak. 
It turns out that the potential for storage events to occur when 
$E$ and $C$ are both higher than average (the cases 1, above); versus the potential  
when $E$ and $C$ are
both lower than average (the cases 2) are not necessarily equal -- storage potential may differ
across the spectrum from 1 to 2.
ATA effects quantify the extent to which asymmetries of association between $E$ and 
$C$ are aligned with differences in the potential for storage across the spectrum from 1 to 2.
If $E$ and $C$ are largely unassociated in the parts of their distribution where
storage potential is high, ATA effects will be high. ATA effects quantify how temporal niche partitioning
varies across $EC$ spectrum spanned by the cases 1-2 sketched above.

%Goal para
We previously demonstrated that ATAs are common in nature and can influence ecological phenomena
\citep{Ghosh_2020_AER, Ghosh_2020_Eco, Ghosh_2020_ecolevol, Ghosh_2021}; and our arguments above
indicate that ATAs may influence coexistence. Thus, we here seek to quantify the consequences 
of ATAs for competition and coexistence, via the following specific goals. 
G1) We augment MCT to show formally how ATAs play a role in coexistence. 
MCT decomposes a measure of coexistence into components due to each of several mechanisms. 
We address how ATAs contribute to these components. 
G2) We explore, using the lottery model, the relative magnitudes 
of the contributions to coexistence that come
from ATAs compared to other, previously known mechanisms of coexistence; and we determine whether species may 
sometimes coexist with ATAs but not without, or vice versa. This is to help 
illuminate whether ATA contributions to coexistence 
are likely to be negligible or meaningfully large. 
G3) We provide an empirical example of species coexistence which highlights the role ATAs can play and the
potential importance of ATAs.
The example is a laboratory diatom system. In the Discussion, we consider whether climate change may alter
ATAs of environmental variables and thereby influence competition.
Overall, our study presents a new mechanism of species coexistence
and a means of understanding its importance.

@@

We here pursue goal G1 from the Introduction: to augment MCT to show the role of ATAs for coexistence.

@@

We first develop our new theory for a general two-species model.
Our notation parallels
that of \cite{Ellner_2016,Ellner_2019}. It is assumed that the growth rate $r_i(t)=\ln(N_i(t+1)/N_i(t))$ 
of species $i$ can be written as an increasing function of
an environmental factor $E_i(t)$ and as a decreasing function of 
competitive pressure on species $i$, $C_i(t)$, i.e., $r_i(E_i,C_i)$. We also assume, for simplicity, 
that the $E_i(t)$ are iid through time, and likewise for the
$C_i(t)$. Though the iid assumption is probably unnecessary, it simplifies the
presentation of new ideas.

@@

%Climate change may alter ATAs of coexistence-relevant ATAs
An additional motivation for quantifying the importance of ATAs for species coexistence 
is that there is reason to believe climate change may
alter ATAs of environmental variables which impact coexistence, as follows.
It is well known that climate change can alter the intensity of extreme environmental events, including
heat waves, floods, hurricanes and fires \citep{Meehl_2004, Jentsch_2007, Ummenhofer_2017, Lyon2019, Keelings2020}.
If extreme events are not only becoming more intense, but are also increasing in duration (e.g., heat waves 
are not only hotter but also last longer, \citealp{Meehl_2004,Lyon2019,Keelings2020}), 
then the nature of the covariation between phenologically separated environmental signals
may be modified by these changes; in particular, ATAs between such variables may be altered.
For instance, imagine a case in which early spring temperature influences the growth of a plant species, A,
whereas late spring temperature influences plant species B. If these species are competitors, then EC covariation
contributing to their coexistence, or failure to coexist, could boil down to covariation
between early and late spring temperatures (e.g., if competitive dynamics are governed by the lottery model with $E_i$ representing
early spring temperature and $E_j$ representing late spring temperature, see previous paragraph). 
Because longer lasting heat waves are more likely to contribute
to both early and late spring temperature, a tendency for heat waves to become both more extreme and
longer lasting increases the degree of right-tail association between early and late spring temperature.
Likewise, a tendency for cold snaps to become both more intense (i.e., colder temperatures) and longer lasting can increase
left-tail association between early- and late-spring temperatures. As we showed in this study, changes in ATAs can then
result in different competitive outcomes between species.
To our knowledge, the potential importance of climatic changes in ATAs for species coexistence has
not been studied, though the importance for coexistence of changes
in means and variability of environmental variables has been studied in many systems
[e.g., \cite{White_2001, Adler_2006, Jentsch_2007, Angert_2009, diPaola_2012, Descombes_2020, Usinowicz_2021}].
Changes in variability and in extreme events are distinct from the changes in ATAs we
consider here, because ATAs pertain to \emph{relationships} between environmental and other
variables in the extremes, which are distinct from and can be altered independently
of changes in the extremes of the univariate marginal distributions themselves \citep{Nelsen2006,Ghosh_2020_AER}.
Additional discussion points are in SI sections \ref{SI-SIsect:ATA_specialists}-\ref{SI-sect:storage_importance}.

### 2023-07-31 paper/SupMat.Rnw
working on brevity and clarity throughout. also added Jasmin's water limitation example to the Intro

>>>>

\section{   ATA specialists}\label{SIsect:ATA_specialists}

Contributions of ATAs to species coexistence open the possibility that
``ATA specialists'' may exist. For instance, if a strong competitor is present across a
collection of habitat patches on a landscape and the GWR of a weaker competitor is only positive 
in a patch if there are 
ATAs between the temporal fluctuations of two environmental variables in that patch,
then the weaker competitor can only persist in patches for which ATAs are present.
The weaker competitor can therefore perhaps be regarded as an ``ATA specialist'' in the landscape. EC-associations
can boil down to associations between two environmental variables, as was the case
for the lottery model. For the lottery model, $B_i(t)=E_i(t)$ was assumed, and $C_{i \bs i}$ simplified to $E_j(t)/\delta$ (see SI section 
\ref{SIsect:TheoryForLotteryBoth}). Thus, EC-associations parallel associations between the two
environmental variables $E_i(t)$ and $E_j(t)$ for the lottery model, and ATA specialists may exist. 
Mechanisms of coexistence should correspond to
means by which niche differentiation is possible. Thinking about our new mechanism of 
coexistence in terms of possible ATA specialists helps highlight this connection. Likewise,
because we have described ATA exclusion as well as ATA rescue, it is possible to 
imagine a scenario in which a weaker competitor can only exist in habitat patches 
for which ATAs between environmental variables are absent, and is hence a specialist on scenarios with symmetric
tail associations of variables (an ``STA specialist''). Additional 
research will be needed to assess to what degree these possibilities are realized 
in nature.

\section{   Differences with the original, analytic version of MCT}\label{SIsect:SimsVAnalysis}

Our approach to exploring the influence of ATAs on coexistence relies on the computational
extension of MCT of \cite{Ellner_2019},
and our results help demonstrate strengths of that approach.
Storage effects as computed using the original, analytic approach to MCT 
\citep{Chesson_1994,Chesson_2000} should equal our term $\Delta^{[E \nat C]}$, 
which we refer to as the contributions to coexistence of correlation \emph{per se}.
This is because the weak-noise/Taylor approximation approach used in the original 
MCT neglects the effects of ATAs when higher-order terms
in Taylor expansions are dropped. The term $\Delta^{(E \shp C)}$, which our theory 
inherits from \cite{Ellner_2016}, is also dropped. These observations do not diminish 
the major strengths of the 
original, analytic approach, which are well known and also elaborated by \cite{Ellner_2019};
one such strength is the generality that analytic approaches provide.
It may be possible to expand the mathematics of the original MCT to consider ATAs.
